---
title: "Data Exploration with CHiMiC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Exploration with CHiMiC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# 1. Introduction
This vignette demonstrates how to perform Exploratory Data Analysis (EDA) on mixed-type clinical data after applying the CHiMiC (Causal High-dimensional Mixed-type Clustering) function. Our goal is to visualize the clustering results, the conditional average treatment effects (TE), and the informative high-dimensional features selected by the model.

# 2. Simulating Data and Fitting the Model
First, we simulate a dataset containing both continuous and categorical clinical covariates, along with a simulated treatment effect. Then, we fit the CHiMiC model.

```{r, warning = FALSE, message = FALSE}
library(CHiMiC)

set.seed(123)
simdata <- simData(n = 1000, s1 = 20, s2 = 5, ratio = 0.2)

Z <- simdata$Z
G <- simdata$G
q1 <- simdata$q1
q2 <- simdata$q2
cate <- simdata$CATE
w <- 0.5
lambda1 <- 50
lambda2 <- 50

# For demonstration purposes, max_iter and inner_iter are set to small values.
# In practice, larger iterations (e.g., 100) are needed for precise estimation.
chimic_obj <- CHiMiC(Z, q1, q2, TE = cate, w, G, lambda1, lambda2,
                     max_iter = 50, inner_iter = 50, tol = 1e-2)
```

# 3. Visualizing the Treatment Effect (TE)
Since CHiMiC performs TE-guided clustering, it is crucial to verify that the identified subgroups exhibit distinct responses to the treatment. We can visualize this using a boxplot overlaid with jittered points.
```{r, warning = FALSE, message = FALSE}
# Install packages if not already installed: install.packages(c("ggplot2", "dplyr", "tidyr", "patchwork"))
library(ggplot2)
library(dplyr)
library(tidyr)

# Extract clustering results and convert to a factor variable
cluster_assign <- as.factor(chimic_obj$clusters)

plot_df <- data.frame(
  Cluster = cluster_assign,
  TE = cate # Treatment effect
)

# Distribution of Treatment Effect (TE) across different clusters
p_te <- ggplot(plot_df, aes(x = Cluster, y = TE, fill = Cluster)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1.5, color = "darkgray") +
  theme_minimal() +
  labs(title = "A. Treatment Effect (TE) by Cluster",
       y = "Conditional Average Treatment Effect",
       x = "Identified Clusters") +
  theme(legend.position = "none")

print(p_te)
```

# 4. Exploring Selected High-Dimensional Features
CHiMiC incorporates an MCP-based variable selection method. To display the selected informative continuous and categorical variables across different clusters, a complex heatmap is highly recommended.
```{r, fig.width = 8, fig.height = 8, warning = FALSE, message = FALSE}
# Install ComplexHeatmap if not already installed: BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize) # For generating continuous color mapping

# Extract clustering results as character for annotation
clusters <- as.character(chimic_obj$clusters)

# Extract selected continuous and categorical variables
Z_cont_sel <- Z[, chimic_obj$selected_features1, drop = FALSE]
Z_cat_sel  <- Z[, q1 + chimic_obj$selected_features2, drop = FALSE]

# Convert categorical variables to factor/character for discrete color mapping in heatmap
Z_cat_sel <- as.data.frame(Z_cat_sel)
colnames(Z_cat_sel) <- paste0("Cat_Var_", chimic_obj$selected_features2)
for(i in 1:ncol(Z_cat_sel)) {
  Z_cat_sel[, i] <- as.character(Z_cat_sel[, i])
}

# Get the order of samples based on clustering results
order_idx <- order(chimic_obj$clusters)

# Sort all data based on the cluster order
sorted_clusters <- clusters[order_idx]
sorted_Z_cat    <- Z_cat_sel[order_idx, , drop = FALSE]

# Scale continuous variables (column-wise standardization), then transpose 
# (Heatmaps typically represent variables as rows and samples as columns)
scaled_Z_cont <- scale(Z_cont_sel)
sorted_Z_cont <- t(scaled_Z_cont[order_idx, , drop = FALSE])
rownames(sorted_Z_cont) <- paste0("Cont_Var_", chimic_obj$selected_features1)

# Create top annotation containing Clusters and categorical variables
anno_df <- data.frame(Cluster = sorted_clusters)
anno_df <- cbind(anno_df, sorted_Z_cat)

top_anno <- HeatmapAnnotation(
  df = anno_df,
  show_annotation_name = TRUE,
  col = list(Cluster = c("1" = "#E41A1C", "2" = "#377EB8", "3" = "#4DAF4A"))
)

# Define color mapping range for continuous variables (e.g., -2 to 2 standard deviations)
col_fun <- colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

ht <- Heatmap(
  sorted_Z_cont,
  name = "Scaled\nValue",        
  col = col_fun,                 
  top_annotation = top_anno,     
  cluster_columns = FALSE,       
  cluster_rows = TRUE,           
  column_split = sorted_clusters,  
  show_column_names = FALSE,     
  row_names_side = "left",       
  row_title = "Selected Continuous Variables",
  column_title = "Samples Ordered by Clusters"
)

# Draw the heatmap
draw(ht, merge_legend = TRUE)
```

# 5. Detailed Feature Comparison
We can also inspect individual features to understand how specific covariates drive the cluster assignments.

```{r, warning = FALSE, message = FALSE}
# For demonstration, extract the first continuous variables
plot_df$Cont_V1 <- Z[, 1]

# Extract the first categorical variable (column index is q1 + 1) and convert to factor
plot_df$Cat_V1 <- as.factor(Z[, q1 + 1])

# Comparison of a continuous variable across different clusters (Boxplot)
p_cont <- ggplot(plot_df, aes(x = Cluster, y = Cont_V1, fill = Cluster)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  labs(title = "B. Continuous Variable 1 by Cluster",
       y = "Value of Cont_V1",
       x = "Identified Clusters") +
  theme(legend.position = "none")

# Proportions of a categorical variable across different clusters (100% Stacked Bar Chart)
p_cat <- ggplot(plot_df, aes(x = Cluster, fill = Cat_V1)) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_y_continuous(labels = scales::percent_format()) + # Display Y-axis as percentages
  theme_minimal() +
  labs(title = "C. Categorical Variable 1 Proportions",
       y = "Proportion",
       fill = "Category")

library(patchwork) 
final_plot <- p_cont | p_cat
print(final_plot)
```


